---
title: "Statistical Analysis of Temporal and Spatial Trends in US Covid-19 Cases and Deaths"
author: "Jason Gong and Micah Swann"
affiliation: "UC Davis"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,warning=FALSE,message=FALSE, include = FALSE}
## REQUIRED PACKAGES
# install.packages("gridExtra")
# packages = c("gtrendsR","tidyverse","usmap")
# install.packages("mapproj")
# importing libraries

## REQUIRED LIBRARIES
 library(dplyr)
 library(tidyverse)
 library(data.table)
 library(ggplot2)  
 library(scales)
 library(gridExtra)
 library(lmtest)
 library(forecast)
```

# Introduction

Covid-19 is a novel, highly contagious, acute respiratory virus that was first identified in December 2019 in Wuhan, China. Over the course of the following 14 months, this virus spread rapidly to every corner of the globe, becoming one of the deadliest pandemics in recorded history. In the United States, the first confirmed Covid-19 case was identified in January 2020 and by mid-March there were confirmed cases in every single state and North American territory. In the midst of this rapid pandemic spread, epidemiologists and modelers struggled to accurately forecast the spatial and temporal trends in cases and deaths. However, with regularly updated, publicly-available covid tracking data, a sufficient amount of data now exists to retroactively examine how cases and deaths evolved over the course of this 14 month period. This study utilizes the New York Times Covid Tracking Data to statistically analyze trends in the timeseries of Covid-19 cases and deaths as well as the spatial development of cases at the state level across the united states using cluster analysis. 

# Data Description

## Data Sources

Due to the fragmented nature of the US public health system, there is no centralized governmental data repository that is updated daily with Covid-19 case and death data. Instead, this study obtained data from the New York Times (NYTimes) Covid-19 Tracking Project (https://github.com/nytimes/covid-19-data). The NYTimes relies on dozens of reporters across multiple time zones to regularly update this tracking database with new information from press conferences, report releases, and local databases. Datasets utilized in this analysis reported the daily cumulative case and death counts in the US aggregated at the national, state and county level (US.csv, US-states.csv, US-counties.csv), respectively.  Demographic data on state populations were also obtained from the US census bureau to compare per capita rates.

## Data Formatting

All data analysis and visualization for this study was conducted in RStudio. The dataset was filtered to only examine cases and deaths reported from the beginning of March 2020 through the middle of February 2021. Raw data was reported as cumulative cases and deaths through time. To examine daily statistics, a filtering function was applied to calculate the finite difference between each consecutive reporting day.

# Exploratory Data Analysis

## Timeseries Visualization

An initial exploratory data analysis (EDA) was conducted to both elucidate trends and characteristics of the dataset and to guide the model development process. To better understand the temporal evolution of daily new cases and deaths in the US, a timeseries for both of these parameters was first generated (Figure 1).  The timeseries of daily US Covid-19 cases depicts four distinct regimes in the change in daily covid 19 cases throughout the course of the pandemic. From March through the end of May 2020, the number of cases grew logistically; growing exponentially in March before asymptoting at a maximum daily new case load of 25,000-30,000 individuals through April and May. Similar but larger magnitude growth trends are evident in June through August, asymptoting around 65,000 daily new cases, and October through December 2020, asymptoting around 250,000 daily new cases. Note that the sharp drop in cases around the end of December is likely a reflection of a decline in reporting around the winter holidays and not a reflection of the actual drop in the real case load. From January 2021 onwards, the case trend differs from the early regimes, with a noticeable linear decline in the reported case numbers through December. Another interesting aspect of this dataset is the seven-day oscillation in the case numbers. New case numbers always tend to be lower on Saturdays and Sundays than during weekdays, reflecting the fact that many labs do not report case numbers on the weekend. The timeseries of daily Covid-19 deaths shows a similar logistic growth rate to the case rate in the early spring 2020. However, the number of deaths, drops significantly from mid-May 2020 and oscillates around 1000 cases a days until November 2020 when the number of daily new deaths rises again, fluctuating around 3000 deaths per days. The seven oscillations, observed in the timeseries of new cases, is even more prominent in the death data, with significant drops in reported deaths during the weekend. This initial visualization and review makes evident that that there are clear similarities and differences in the functional trends between both datasets.\  
\  

```{r, echo = FALSE}
# read in NYtimes Covid data tables 

# Cumulative Daily Cases and Deaths by 
# 1 - County Level
cty_data <-read.csv("us-counties.csv") 

# 2 - State Level
state_data <-read.csv("us-states.csv")
state_pop <-read.csv("state_pop.csv")

# 3 - 
us_data <-read.csv("us.csv")

# transforming date column from CHARACTER to DATE class
cty_data$date <- as.Date(cty_data$date)
state_data$date <- as.Date(state_data$date,format="%Y-%m-%d")
us_data$date <- as.Date(us_data$date)

# filter out dates before March 1st, 2020
cty_data <- base::subset(cty_data,date>"2020-02-29")
state_data <- base::subset(state_data,date>"2020-02-29")
us_data <- base::subset(us_data,date>"2020-02-29")

# Examination of US-wide Temporal Trends

# Use `filter()` to add a column named `new_cases` and 'new_deaths'
us_data$new_cases = as.numeric(stats::filter(us_data$cases,filter=c(1,-1), sides=1))

us_data$new_deaths = as.numeric(stats::filter(us_data$deaths,filter=c(1,-1), sides=1))
 
# Change first entry fo new deaths and new cases from NA to 0
us_data$new_cases[1]<-0
us_data$new_deaths[1]<-0

# Add Data on Season
season <- c("Spring","Summer","Winter","Fall")

# create df for seasons
season_df = data.frame(c(1:12),c("Winter","Winter","Spring","Spring","Spring","Summer","Summer","Summer","Fall","Fall","Fall","Winter"))
names(season_df)[1] <- "month"
names(season_df)[2] <- "season"

for (i in 1:length(us_data$date))
{
  if (month(us_data$date[i]) >= 3 & month(us_data$date[i]) < 6) {
    us_data$season[i] = season[1]
  } else if (month(us_data$date[i]) >= 6 & month(us_data$date[i]) < 9) {
    us_data$season[i] = season[2]
  } else if (month(us_data$date[i]) >= 9 & month(us_data$date[i]) < 12) {
    us_data$season[i] = season[3]
  } else {
    us_data$season[i] = season[4]
  }
}
us_data$season = as.factor(us_data$season)

## Plotting Below

# Timeseries of Covid Cases and Deaths through time
# Value used to transform the data
p1 <- ggplot(us_data, aes(x = date, y = new_cases)) +
            geom_line(color = "darkred") +
  labs(title="US Covid-19 Cases and Deaths",  y="Daily New Cases", x="")+
  theme(plot.title = element_text(hjust = 0.5))
  

p2 <- ggplot(us_data, aes(x = date, y = new_deaths)) +
            geom_line(color = "black") +
  labs(  y="Daily New Deaths", x="Date")
   
grid.arrange(p1,p2,nrow=2)
```
\begin{center}
Figure 1 - Timeseries of daily new Covid-19 cases (top) and deaths (bottom) in the United States, March 2020 - February 2021.
\end{center}

## Case-Death Scatter Plot

To further examine how the relationship between daily new US covid cases and deaths changed through time, a scatter plot of these two variables was generated (Figure 2). Having previously identified four distinct regimes in the case growth rate, these points were colored by the season for each observation. The scatter plot further highlights the different trends in the daily case to death ratios during each of these time periods. In the spring 2020, there is an significantly evident positive correlation in the case to death ratio. The points in this season are all closely clustered with the deaths growing exponentially with cases. In the summer 2020 data, there is no clear positive or negative trend in the correlation between the two parameters with the points scatter roughly in a circle. In the fall season, again a positive correlation is evident, however the case to death ratio is four to five times that observed in the spring. Finally in the winter (December 2020 – February 2021), the ratio of cases to deaths decreases but is still positive.   
  
```{r,echo=FALSE}
# scatter plot comparison of Cases Vs Deaths
gg <- ggplot(us_data,aes(x = new_cases, y = new_deaths, colour = season)) +
  geom_point() +
   labs(title="Cases Vs Deaths (US-wide)",y="Daily Deaths", x="Daily Cases")+
  theme(plot.title = element_text(hjust = 0.5))

plot(gg)
```
\begin{center}
Figure 2 - Scatterplot of daily new Covid-19 cases vs deaths in United States. Point colors indicate season
\end{center}

## State Case Rate Box Plots
Description of box plots blah blah blah blah blah
```{r, echo = FALSE, fig.align="center"}
## Examining State Spatial Trends

# 1 - Formatting Data

# remove leading character in State_Pop$State
state_pop$State <- substring(state_pop$State,2)

# remove commas from population and convert to numeric
state_pop$Population <- as.numeric(gsub(",","",state_pop$Population))


# add state population data
state_data <- merge(state_data, state_pop, by.x="state", by.y="State")


# get population density
# remove DC from state_pop list
state_pop <-state_pop [-c(9), ] 
state.density = state_pop$Population/state.area

## get state regional data
states_info <- data.frame(state.name,state.region,state.density)
names(states_info)[1] <- "Name"
names(states_info)[2] <- "Region"
names(states_info)[3] <- "Density"

# add state region data
state_data <- merge(state_data, states_info, by.x="state", by.y="Name")

# categorize into density groups

for (i in 1:length(state_data$Density))
{
  if (state_data$Density[i] <= 50) {
     state_data$Density2[i] <- "<=50"
  } else if (state_data$Density[i] >  50 & state_data$Density[i] <= 100) {
    state_data$Density2[i] <- "50-100"
  } else if (state_data$Density[i] >  100 & state_data$Density[i] <= 400) {
    state_data$Density2[i] <- "100-400"
  } else {
    state_data$Density2[i] <- ">400"
  }
}
state_data$Density2 <- factor(state_data$Density2,
    levels = c('<=50','50-100','100-400','>400'),ordered = TRUE)

# sort by date
state_data<-state_data[order((state_data$date)),]

# Find unique fips values to find unique counties
fips_vect <- unique(cty_data$fips)

# Find unique dates
dates_vect <- unique(cty_data$date)

# find unique states (Exclude territories)
states_vect <-state.name;

# Find number of daily new cases and deaths by state
i <-1

# looping through unique state names
for (val in states_vect)
{
  # subset by state
  dum <- subset(state_data, state == val)
  
  #Use `filter()` to add a column named `new_cases` 
  dum$new_cases = as.numeric(stats::filter(dum$cases,filter=c(1,-1), sides=1))

  # Set Na values to 0. These occur on the first date of obervations
  dum$new_cases[is.na(dum$new_cases)] <- 0
  
  #Use `filter()` to add a column named `new_deaths` 
  dum$new_deaths = as.numeric(stats::filter(dum$deaths,filter=c(1,-1), sides=1))
  
    # Set NaN values to 0. These occur on the first date of obervations
  dum$new_deaths[is.na(dum$new_deaths)] <- 0
  
  # append to dataframe
  if (i>1){
    df1 <- rbind(df1,dum)
  }else{
    df1 <- dum
  }
  i <- i+1
  rm(dum)
}
# final new data frame
state_data2 <- df1

#calculate per capita case and death rates
state_data2$pc_new_cases <- state_data2$new_cases / state_data2$Population
state_data2$pc_new_deaths <- state_data2$new_deaths / state_data2$Population

# extracting year and month
month <- data.frame(format(state_data2$date,format="%m/%y"))
# month <- data.frame(format(state_data2$date,format="%m"))

# add as columns to df
# state_data2<-bind_cols(state_data2,year)
state_data2<-bind_cols(state_data2,month)

# sorting columns
state_data3 <- state_data2[,c(1,14,2:13)]

# renaming columns
names(state_data3)[2] <- "month"

# Calculate pc avg cases for each state
state_mnthly_cases = as.data.table(state_data3)[, mean(pc_new_cases), by = .(month,state, Region,Density2)]
names(state_mnthly_cases)[5] <- "avg_pc_cases"
state_mnthly_cases$Density2 = as.factor(state_mnthly_cases$Density2)
names(state_mnthly_cases)[4] <- "Density"

# Calculate avg deaths for each state
state_mnthly_deaths = as.data.table(state_data3)[, mean(pc_new_deaths), by = .(month,state,Region,Density2)]
names(state_mnthly_deaths)[5] <- "avg_pc_deaths"
state_mnthly_deaths$Density2 = as.factor(state_mnthly_deaths$Density2)
names(state_mnthly_deaths)[4] <- "Density"

# Boxplot of Monthly Cumulative Cases by State

# getting dates in order
dum = unique(state_mnthly_cases$month)
state_mnthly_cases$month <- factor(state_mnthly_cases$month , levels=dum)
state_mnthly_deaths$month <- factor(state_mnthly_deaths$month , levels=dum)

# plotting below
```

```{r, echo = FALSE,fig.width=7,fig.height=4}
# boxplot of cases
mnthly_case_bplot = ggplot(state_mnthly_cases,aes(month, avg_pc_cases)) + 
geom_boxplot() +
  theme(plot.title = element_text(hjust = 0.5))
mnthly_case_bplot + labs(y = "Mean Per Capita Daily Cases by State",x = "Month",
                   title = "Box Plot of Monthly Per Capita Covid-19 Cases by State") 
```
\begin{center}
Figure 3 - Boxplot of daily average new Covid-19 cases by state across US. Data is binned by month
\vspace{5mm} %5mm vertical space
\end{center}
```{r, echo = FALSE,fig.width=7,fig.height=4}
# boxplot of deaths
mnthly_death_bplot = ggplot(state_mnthly_deaths,aes(month, avg_pc_deaths)) + geom_boxplot() +
  theme(plot.title = element_text(hjust = 0.5))
mnthly_death_bplot + labs(y = "Mean Per Capita Daily Deaths by State",x = "Month",
                   title = "Box Plot of Monthly Per Capita Covid-19 Deaths by State")  

```
\begin{center}
Figure 4 - Boxplot of daily average new Covid-19 deaths by state across US. Data is binned by month
\end{center}

## Regional Trends
```{r, echo = FALSE}
# Region Map
all_states <- map_data("state")
all_states$region <- str_to_title(all_states$region)  
stateData <- merge(all_states,states_info,by.x="region",by.y = "Name" )

regionplot <- ggplot()+geom_polygon(data=stateData,aes(x=long, y=lat, group = group, fill=Region),color="grey50")+coord_map()
regionplot
```
\begin{center}
Figure 5 - Map of US States grouped by region \pagebreak
\end{center}

```{r, echo = FALSE,fig.width=7,fig.height=4}
# Color by region
mnthly_case_bplot2 = ggplot(state_mnthly_cases,aes(month, avg_pc_cases)) + 
geom_boxplot(aes(color = Region)) 
mnthly_case_bplot2 + labs(y = "Mean Per Capita Daily Cases by Region",x = "Month",
                   title = "Box Plot of Monthly Per Capita Covid-19 Cases by Region")  
```

\begin{center}
Figure 6 - Boxplot of daily average new Covid-19 cases by state across US. Data is binned by month and categorized by region.
\vspace{5mm} %5mm vertical space
\end{center}

```{r, echo = FALSE,fig.width=7,fig.height=4}
mnthly_death_bplot2 = ggplot(state_mnthly_deaths,aes(month, avg_pc_deaths)) + 
geom_boxplot(aes(color = Region)) 
mnthly_death_bplot2 + labs(y = "Mean Daily Per Capita Deaths by Region",x = "Month",
                   title = "Box Plot of Monthly Per Capita Covid-19 Deaths by Region") 
```

\begin{center}
Figure 7 - Boxplot of daily average new Covid-19 deaths by state across US. Data is binned by month and categorized by region. \\
\end{center}

```{r, echo = FALSE,fig.width=7,fig.height=4}
# Color by density
mnthly_case_bplot3 = ggplot(state_mnthly_cases,aes(month, avg_pc_cases)) + 
geom_boxplot(aes(color = Density)) 
mnthly_case_bplot3 + labs(y = "Mean Per Capita Daily Cases by Density",x = "Month",
                   title = "Box Plot of Monthly Per Capita Covid-19 Cases by Density",fill = "Density [pop/mi^2]")  
```

\begin{center}
Figure 8 - Density Case Boxplot
\vspace{5mm} %5mm vertical space
\end{center}

```{r, echo = FALSE,fig.width=7,fig.height=4}
mnthly_death_bplot3 = ggplot(state_mnthly_deaths,aes(month, avg_pc_deaths)) + 
geom_boxplot(aes(color = Density)) 
mnthly_death_bplot3 + labs(y = "Mean Daily Per Capita Deaths by Density",x = "Month",
                   title = "Box Plot of Monthly Per Capita Covid-19 Deaths by Density",fill = "Density [pop/mi^2]") 
```

\begin{center}
Figure 9 - Density Death Boxplot
\newpage
\end{center}


# Algorithms

## Autoregressive (AR) Forecast Model

Description of AR Model\
Model Development Process\

## Cluster Analysis

\begin{center}
\newpage
\end{center}

# Results

## AR Covid Case Forecast Model
```{r,echo = FALSE,fig.align = 'center'}
# Timeseries Analysis

# Converting to timeseries
new_cases.ts <- ts(us_data$new_cases,frequency = 7)
new_deaths.ts <- ts(us_data$new_deaths,frequency = 7)

## CASES
## plot the ts
#autoplot(new_cases.ts)+
#  xlab("Weeks from from March 1st, 2020") + ylab("Daily New Covid Cases in US") +
# ggtitle("Timeseries of New Covid 19 Cases in US") 

## Decompose
components.ts = decompose(new_cases.ts)
plot(components.ts,xlab = "Weeks since March 1st, 2020")
```
\begin{center}
Figure 10 - Decomposition of Covid Case Data
\vspace{5mm} %5mm vertical space
\end{center}

### Test for Stationarity
```{r}
# Test for Stationarity
tseries::adf.test(new_cases.ts)
```
It is not stationary!\

```{r,echo = FALSE,fig.align = 'center'}
## Remove Non Stationrity through differencing
diff_new_cases.ts <- diff(new_cases.ts)

components.ts2 = decompose(diff_new_cases.ts)
plot(components.ts2,xlab = "Weeks since March 1st, 2020")
```
\begin{center}
Figure 10 - Decomposition of Difference Covid Case Data
\vspace{5mm} %5mm vertical space
\end{center}

```{r}
# Test for Stationarity with differencing
tseries::adf.test(diff(new_cases.ts))
```
This data is stationary!\
```{r,echo = FALSE}
## examine autocorrelation function
new_cases.ts %>% diff() %>% ggtsdisplay(main="")
```
\begin{center}
Figure 11 - Autocorelation function and Partial Autocorrelation Function for difference covid case data
\vspace{5mm} %5mm vertical space
\end{center}

```{r,echo=FALSE, results="hide"}
## Calculate optimal coefficient for model
auto.arima(new_cases.ts, trace=TRUE) 

## Determining AR model
fitARIMA_cases <- arima(new_cases.ts, order=c(0,1,1),seasonal = list(order = c(0,0,2), period = 7),method="ML")
```
### Checking Residuals
```{r,echo=FALSE, results="hide"}

## Check residuals
checkresiduals(fitARIMA_cases)
```
\begin{center}
Figure 11 - Autocorelation function and Partial Autocorrelation Function for difference covid case data
\vspace{5mm} %5mm vertical space
\end{center}

### Visualizing Forecast
```{r,echo=FALSE,fig.width=7,fig.height=4,fig.align='center'}
## plot forecast
autoplot(forecast(fitARIMA_cases))+
  xlab("Weeks from from March 1st, 2020") + ylab("Daily New Covid Cases in US") +
  ggtitle("AR Lag 7 Model for Covid Cases")
```
\begin{center}
Figure 12 - 30 Day forecast for Covid-19 cases in US using AR Lag-7 Model. Dark blue indicates model results, light blue bars indicate 95% confidence interval.
\vspace{5mm} %5mm vertical space
\end{center}

## AR Covid Death Forecast Model
```{r,echo=FALSE}
## Decompose
components.ts = decompose(new_deaths.ts)
plot(components.ts,xlab = "Weeks since March 1st, 2020")
```

```{r}
# Test for Stationarity
tseries::adf.test(new_deaths.ts)
```

```{r,echo=FALSE}
## Remove Non Stationrity through differencing
diff_new_deaths.ts <- diff(new_deaths.ts)

components.ts2 = decompose(diff_new_deaths.ts)
plot(components.ts2,xlab = "Weeks since March 1st, 2020")
```

```{r}
# using differencing method
tseries::adf.test(diff(new_deaths.ts))
```

```{r,echo=FALSE, results="hide"}
## examine autocorrelation function
new_deaths.ts %>% diff() %>% ggtsdisplay(main="")

## Calculate optimal coefficient for model
auto.arima(new_deaths.ts, trace=TRUE) 

## Determining AR model
fitARIMA_deaths <- arima(new_deaths.ts, order=c(1,0,2),seasonal = list(order = c(0,1,1), period = 7),method="ML")
```

```{r,echo=FALSE, results="hide"}
## Check residuals
checkresiduals(fitARIMA_deaths)
```

```{r,echo=FALSE,fig.width=5,fig.height=4,fig.align='center'}
## plot forecast
autoplot(forecast(fitARIMA_deaths))+
  xlab("Weeks from from March 1st, 2020") + ylab("Daily New Covid Deaths in US") +
  ggtitle("AR Lag 7 Model for Covid Deaths")
```

# Conclusions