---
title: "Statistical Analysis of Temporal and Spatial Trends in US Covid-19 Cases and Deaths"
author: "Jason Gong and Micah Swann"
output: ioslides_presentation
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r,warning=FALSE,message=FALSE, include = FALSE}
## REQUIRED PACKAGES
# install.packages("gridExtra")
# install.packages("mapproj")
# install.packages("choroplethr")
# importing libraries
## REQUIRED LIBRARIES
 library(dplyr)
 library(tidyverse)
 library(data.table)
 library(ggplot2)  
 library(scales)
 library(gridExtra)
 library(lmtest)
 library(forecast)
 library(dtwclust)
 library(ggpubr)
 library(gplots)
 library(RColorBrewer)
 library(nnet)
 library(summarytools)
 library(lmtest)
```

## Outline
- Motivation
- EDA
- Timeseries Analysis
    - ARIMA Model
- Spatial Analysis
    - Cluster Analysis

## Motivation
```{r,echo=FALSE}
# read in NYtimes Covid data tables 

# Cumulative Daily Cases and Deaths by 
# 1 - County Level
cty_data <-read.csv("us-counties.csv") 

# 2 - State Level
state_data <-read.csv("us-states.csv")
state_pop <-read.csv("state_pop.csv")

# 3 - 
us_data <-read.csv("us.csv")

# transforming date column from CHARACTER to DATE class
cty_data$date <- as.Date(cty_data$date)
state_data$date <- as.Date(state_data$date,format="%Y-%m-%d")
us_data$date <- as.Date(us_data$date)

# filter out dates before March 1st, 2020
cty_data <- base::subset(cty_data,date>"2020-02-29")
state_data <- base::subset(state_data,date>"2020-02-29")
us_data <- base::subset(us_data,date>"2020-02-29")

# Examination of US-wide Temporal Trends

# Use `filter()` to add a column named `new_cases` and 'new_deaths'
us_data$new_cases = as.numeric(stats::filter(us_data$cases,filter=c(1,-1), sides=1))

us_data$new_deaths = as.numeric(stats::filter(us_data$deaths,filter=c(1,-1), sides=1))
 
# Change first entry fo new deaths and new cases from NA to 0
us_data$new_cases[1]<-0
us_data$new_deaths[1]<-0

# Add Data on Season
season <- c("Spring","Summer","Winter","Fall")

# create df for seasons
season_df = data.frame(c(1:12),c("Winter","Winter","Spring","Spring","Spring","Summer","Summer","Summer","Fall","Fall","Fall","Winter"))
names(season_df)[1] <- "month"
names(season_df)[2] <- "season"

for (i in 1:length(us_data$date))
{
  if (month(us_data$date[i]) >= 3 & month(us_data$date[i]) < 6) {
    us_data$season[i] = season[1]
  } else if (month(us_data$date[i]) >= 6 & month(us_data$date[i]) < 9) {
    us_data$season[i] = season[2]
  } else if (month(us_data$date[i]) >= 9 & month(us_data$date[i]) < 12) {
    us_data$season[i] = season[3]
  } else {
    us_data$season[i] = season[4]
  }
}
us_data$season = as.factor(us_data$season)
```

```{r, echo = FALSE}
## Plotting Below

# Timeseries of Covid Cases and Deaths through time
# Value used to transform the data
p1 <- ggplot(us_data, aes(x = date, y = new_cases)) +
            geom_line(color = "darkred") +
  labs(title="US Covid-19 Cases and Deaths",  y="Daily New Cases", x="")+
  theme(plot.title = element_text(hjust = 0.5))
  

p2 <- ggplot(us_data, aes(x = date, y = new_deaths)) +
            geom_line(color = "black") +
  labs(  y="Daily New Deaths", x="Date")
   
grid.arrange(p1,p2,nrow=2)
```

```{r, echo = FALSE, fig.align="center", fig.width = 7, fig.height =3}
## Examining State Spatial Trends

# 1 - Formatting Data

# remove leading character in State_Pop$State
state_pop$State <- substring(state_pop$State,2)

# remove commas from population and convert to numeric
state_pop$Population <- as.numeric(gsub(",","",state_pop$Population))


# add state population data
state_data <- merge(state_data, state_pop, by.x="state", by.y="State")


# get population density
# remove DC from state_pop list
state_pop <-state_pop [-c(9), ] 
state.density = state_pop$Population/state.area

## get state regional data
states_info <- data.frame(state.name,state.region,state.density)
names(states_info)[1] <- "Name"
names(states_info)[2] <- "Region"
names(states_info)[3] <- "Density"

# add state region data
state_data <- merge(state_data, states_info, by.x="state", by.y="Name")

# categorize into density groups

for (i in 1:length(state_data$Density))
{
  if (state_data$Density[i] <= 50) {
     state_data$Density2[i] <- "<=50"
  } else if (state_data$Density[i] >  50 & state_data$Density[i] <= 100) {
    state_data$Density2[i] <- "50-100"
  } else if (state_data$Density[i] >  100 & state_data$Density[i] <= 400) {
    state_data$Density2[i] <- "100-400"
  } else {
    state_data$Density2[i] <- ">400"
  }
}
state_data$Density2 <- factor(state_data$Density2,
    levels = c('<=50','50-100','100-400','>400'),ordered = TRUE)

# sort by date
state_data<-state_data[order((state_data$date)),]

# Find unique fips values to find unique counties
fips_vect <- unique(cty_data$fips)

# Find unique dates
dates_vect <- unique(cty_data$date)

# find unique states (Exclude territories)
states_vect <-state.name;

# Find number of daily new cases and deaths by state
i <-1

# looping through unique state names
for (val in states_vect)
{
  # subset by state
  dum <- subset(state_data, state == val)
  
  #Use `filter()` to add a column named `new_cases` 
  dum$new_cases = as.numeric(stats::filter(dum$cases,filter=c(1,-1), sides=1))

  # Set Na values to 0. These occur on the first date of obervations
  dum$new_cases[is.na(dum$new_cases)] <- 0
  
  #Use `filter()` to add a column named `new_deaths` 
  dum$new_deaths = as.numeric(stats::filter(dum$deaths,filter=c(1,-1), sides=1))
  
    # Set NaN values to 0. These occur on the first date of obervations
  dum$new_deaths[is.na(dum$new_deaths)] <- 0
  
  # append to dataframe
  if (i>1){
    df1 <- rbind(df1,dum)
  }else{
    df1 <- dum
  }
  i <- i+1
  rm(dum)
}
# final new data frame
state_data2 <- df1

#calculate per capita case and death rates
state_data2$pc_new_cases <- state_data2$new_cases / state_data2$Population
state_data2$pc_new_deaths <- state_data2$new_deaths / state_data2$Population

# extracting year and month
month <- data.frame(format(state_data2$date,format="%m/%y"))
# month <- data.frame(format(state_data2$date,format="%m"))

# add as columns to df
# state_data2<-bind_cols(state_data2,year)
state_data2<-bind_cols(state_data2,month)

# sorting columns
state_data3 <- state_data2[,c(1,14,2:13)]

# renaming columns
names(state_data3)[2] <- "month"

# Calculate pc avg cases for each state
state_mnthly_cases = as.data.table(state_data3)[, mean(pc_new_cases), by = .(month,state, Region,Density2)]
names(state_mnthly_cases)[5] <- "avg_pc_cases"
state_mnthly_cases$Density2 = as.factor(state_mnthly_cases$Density2)
names(state_mnthly_cases)[4] <- "Density"

# Calculate avg deaths for each state
state_mnthly_deaths = as.data.table(state_data3)[, mean(pc_new_deaths), by = .(month,state,Region,Density2)]
names(state_mnthly_deaths)[5] <- "avg_pc_deaths"
state_mnthly_deaths$Density2 = as.factor(state_mnthly_deaths$Density2)
names(state_mnthly_deaths)[4] <- "Density"

# Boxplot of Monthly Cumulative Cases by State

# getting dates in order
dum = unique(state_mnthly_cases$month)
state_mnthly_cases$month <- factor(state_mnthly_cases$month , levels=dum)
state_mnthly_deaths$month <- factor(state_mnthly_deaths$month , levels=dum)

# plotting below
```
## EDA
- Timeseries Visualization of US Covid-19 Data
- Case Vs Death scatter plots
- Boxplots of daily averages binned by month at State Level
  - grouped by region
  - grouped by density
  
## Scatterplot
```{r,echo=FALSE}
# scatter plot comparison of Cases Vs Deaths
gg <- ggplot(us_data,aes(x = new_cases, y = new_deaths, colour = season)) +
  geom_point() +
   labs(title="Cases Vs Deaths",subtitle="US Country-wide Covid Counts",y="Daily Deaths", x="Daily Cases")

plot(gg)
```

## Regional and Demographic Trends
```{r, echo = FALSE}
# Region Map
all_states <- map_data("state")
all_states$region <- str_to_title(all_states$region)  
stateData <- merge(all_states,states_info,by.x="region",by.y = "Name" )

regionplot <- ggplot()+geom_polygon(data=stateData,aes(x=long, y=lat, group = group, fill=Region),color="grey50")+coord_map()
regionplot
```

## Regional Box Plots
```{r, echo = FALSE,fig.width=7,fig.height=2.3}
# Color by region
mnthly_case_bplot2 = ggplot(state_mnthly_cases,aes(month, avg_pc_cases)) + 
geom_boxplot(aes(color = Region)) 
mnthly_case_bplot2 + labs(y = "Mean Per Capita Daily \n Cases by Region",x = "Month",
                   title = "Box Plot of Monthly Per Capita Covid-19 Cases by Region")  
```

```{r, echo = FALSE,fig.width=7,fig.height=2.3}
mnthly_death_bplot2 = ggplot(state_mnthly_deaths,aes(month, avg_pc_deaths)) + 
geom_boxplot(aes(color = Region)) 
mnthly_death_bplot2 + labs(y = "Mean Daily Per Capita \n Deaths by Region",x = "Month",
                   title = "Box Plot of Monthly Per Capita Covid-19 Deaths by Region") 
```

## ARIMA Timeseries Forecasting Model

ARIMA - AutoRegressive Integrated Moving Average Model
$$Y_t = \alpha + \beta_1 Y_{t-1} + \beta_2 Y_{t-2} + \ldots +$$
$$ \beta_p Y_{t-p}\epsilon_p + \epsilon_t + \theta_1 \epsilon_{t-1} + \theta_2 \epsilon_{t-2} + \ldots + $$
$$\theta_q \epsilon_{t-q}$$


## ARIMA Model Development Process
- Decomposition of data to examine trends and oscillatory patterns
- ADF test to check for non-stationarity
- Difference scheme to stationarize data
- Visualization of ACF and PCF to determine lags
- Calculations of Optimal Parameters for seasonal and non-seasonal trends

## Decomposed Raw Covid Case Data
```{r,echo = FALSE,fig.align = 'center'}
# Timeseries Analysis of Cases

# Converting to timeseries
new_cases.ts <- ts(us_data$new_cases,frequency = 7)
new_deaths.ts <- ts(us_data$new_deaths,frequency = 7)

## Decompose
components.ts = decompose(new_cases.ts)
plot(components.ts,xlab = "Weeks since March 1st, 2020")
```

## ADF Test for Stationarity on Raw Data
```{r}
# Test for Stationarity
tseries::adf.test(new_cases.ts)
```

## Decomposed Difference Case Data
```{r,echo = FALSE,fig.align = 'center'}
## Remove Non Stationrity through differencing
diff_new_cases.ts <- diff(new_cases.ts)

components.ts2 = decompose(diff_new_cases.ts)
plot(components.ts2,xlab = "Weeks since March 1st, 2020")
```

## ADF Test for Stationarity on Differenced Data
```{r}
# Test for Stationarity with differencing
tseries::adf.test(diff(new_cases.ts))
```

## ACF and PCF
```{r,echo=FALSE, results="hide"}
## examine autocorrelation function
new_cases.ts %>% diff() %>% ggtsdisplay(main="")

## Calculate optimal coefficient for model
auto.arima(new_cases.ts, trace=TRUE) 

## Determining AR model
fitARIMA_cases <- arima(new_cases.ts, order=c(0,1,1),seasonal = list(order = c(0,0,2), period = 7),method="ML")
```

## Checking Residuals
```{r,echo=FALSE, results="hide"}

## Check residuals
checkresiduals(fitARIMA_cases)

```

```{r,include=FALSE}
## For Deaths
## Decompose
components.ts = decompose(new_deaths.ts)
plot(components.ts,xlab = "Weeks since March 1st, 2020")

# Test for Stationarity
tseries::adf.test(new_deaths.ts)
## Remove Non Stationrity through differencing
diff_new_deaths.ts <- diff(new_deaths.ts)

components.ts2 = decompose(diff_new_deaths.ts)
plot(components.ts2,xlab = "Weeks since March 1st, 2020")

# using differencing method
tseries::adf.test(diff(new_deaths.ts))

## examine autocorrelation function
new_deaths.ts %>% diff() %>% ggtsdisplay(main="")

## Calculate optimal coefficient for model
auto.arima(new_deaths.ts, trace=TRUE) 

## Determining AR model
fitARIMA_deaths <- arima(new_deaths.ts, order=c(1,0,2),seasonal = list(order = c(0,1,1), period = 7),method="ML")
```

## Forecast Model Results
```{r,echo=FALSE,fig.width=7,fig.height=2.5,fig.align='center'}
fmodel_cases <- forecast(fitARIMA_cases)

## plot forecast
autoplot(forecast(fitARIMA_cases))+
  xlab("Weeks from from March 1st, 2020") + ylab("Daily New Covid Cases in US") +
  ggtitle("AR Lag 7 Model for Covid Cases")
```

```{r,echo=FALSE,fig.width=7,fig.height=2.5,fig.align='center'}
## plot forecast

fmodel_death <- forecast(fitARIMA_deaths)
autoplot(forecast(fitARIMA_deaths))+
  xlab("Weeks from from March 1st, 2020") + ylab("Daily New Covid Deaths in US") +
  ggtitle("AR Lag 7 Model for Covid Deaths")
```

## Timeseries Analysis and Conclusions
- ARIMA models were developed to forecast US-wide Covid-19 case and death data for a 2 week period
- both models agreed within 10 percent of observed numbers at end of predication timeframe
- models reproduced long term trends as well as "artificial" oscillations in reporting numbers
- this analysis could be improved by incorporating other parameters into a multivariate regression

# Spatial Trend Anlysis
## Does geographical distance influence the deaths and cases growth?
- Smoothed the time series by 7-days sliding window averaging 
- Applied a K-shape distance algorithm for both cases and deaths time series
- Calculated geographical distances between any two states
- Run a correlation between shape of time series and the geographical distances

## Smoothed time series
```{r make case and dealth matrix, include=FALSE}
state_data2 <- state_data2 %>%
  mutate(avg_new_cases = stats::filter(new_cases, filter=rep(1/7,7), side=1),
         avg_new_deaths = stats::filter(new_deaths, filter=rep(1/7,7), side=1))


ts_avg_case_matrix <- matrix(, nrow = length(states_vect), ncol = length(dates_vect))
ts_avg_death_matrix <- matrix(, nrow = length(states_vect), ncol = length(dates_vect))

for (i in 1:length(states_vect)) {
  for (j in 1:length(dates_vect)) {
    tryCatch({ts_avg_case_matrix[i,j] <- state_data2$avg_new_cases[(state_data2$state == states_vect[i] 
                                                            & state_data2$date == dates_vect[j])]},
                                    error=function(e){ts_avg_case_matrix[i,j] <- NA})
    tryCatch({ts_avg_death_matrix[i,j] <- state_data2$avg_new_deaths[(state_data2$state == states_vect[i] &
                                                             state_data2$date == dates_vect[j])]},
                                    error=function(e){ts_avg_death_matrix[i,j] <- NA})
  }
}

rownames(ts_avg_case_matrix) <- states_vect
rownames(ts_avg_death_matrix) <- states_vect
colnames(ts_avg_case_matrix) <- dates_vect
colnames(ts_avg_death_matrix) <- dates_vect

ts_avg_case_matrix <- ts_avg_case_matrix[ , colSums(is.na(ts_avg_case_matrix)) == 0]
ts_avg_death_matrix <- ts_avg_death_matrix[ , colSums(is.na(ts_avg_death_matrix)) == 0]

```
```{r plot time series data, echo=FALSE, results="hide", message=FALSE, warning = FALSE, fig.width=8, fig.height=4}
cl <- rainbow(50)

par(mfrow=c(1,2)) 
plot(0,0,xlim = c(0,330),ylim = c(0,10),type = "n", main= 'Cases time series data for each state', xlab='Time', ylab='Log num of cases')
for (i in 1:50){
    lines(log(ts_avg_case_matrix[i,]),col = cl[i], type = 'l')
}

plot(0,0,xlim = c(0,330),ylim = c(0,8),type = "n", main='Deaths time series data for each state', xlab='Time', ylab='Log num of deaths')
for (i in 1:50){
    lines(log(ts_avg_death_matrix[i,]),col = cl[i], type = 'l')
}
```

## K-shape Distance Matrix for time series data
```{r SBD distance matrix and geographical distance matrix, echo=FALSE, include=FALSE}
suppressPackageStartupMessages(library(dtw))

dtw_distmat_case <- proxy::dist(ts_avg_case_matrix, method = "SBD", 
                           upper = TRUE, diag = TRUE)
dtw_distmat_death <- proxy::dist(ts_avg_death_matrix, method = "SBD", 
                           upper = TRUE, diag = TRUE)

dtw_distmat_case <- as.matrix(dtw_distmat_case)
dtw_distmat_death <- as.matrix(dtw_distmat_death)
us_distance_matrix <- as.matrix(dist(cbind(state.center$x, state.center$y)))
rownames(us_distance_matrix) <- rownames(dtw_distmat_death)
colnames(us_distance_matrix) <- colnames(dtw_distmat_death)
df <- data.frame(case=dtw_distmat_case[lower.tri(dtw_distmat_case, diag=FALSE)],
                 death=dtw_distmat_death[lower.tri(dtw_distmat_death, diag=FALSE)],
                 geo_dis=us_distance_matrix[lower.tri(us_distance_matrix)])
```
```{r echo=FALSE, fig.width=4, fig.height=3}
heatmap(dtw_distmat_case, Colv = NA, Rowv = NA, symm =T,  cex.main=0.8, margins = c(3, 3), 
        main = 'K-shape distance cases', cexCol = 0.5, cexRow = 0.5)
heatmap(dtw_distmat_death, Colv = NA, Rowv = NA, symm =T, margins = c(3, 3), 
        main = 'K-shape distance deaths', cexCol = 0.5, cexRow = 0.5)

```

## Geographical distances between states
```{r echo=FALSE, fig.width=4, fig.height=3, fig.align = 'center'}
heatmap(us_distance_matrix, Colv = NA, Rowv = NA, symm =T, margins = c(3, 3), 
        main = 'Euclidean Geo distance', cexCol = 0.5, cexRow = 0.5)
```

## Correlation between geographical distances and case shape
```{r Correlation test for distances cases, echo=FALSE}
cor.test(df$case, df$geo_dis, method = "pearson")
```

## Correlation between geographical distances and death shape
```{r Correlation test for distances, echo=FALSE}
cor.test(df$death, df$geo_dis, method = "pearson")
```

## Correlation between geographical distances and case/death shape
```{r Correlation plot, fig.width=4, fig.height=4, echo=FALSE, results='hide', message=FALSE}
ggscatter(df, x = "case", y = "geo_dis", size = 0.8,
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "geo distance", ylab = "case K-shape distance")

ggscatter(df, x = "death", y = "geo_dis", size = 0.8,
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "geo distance", ylab = "death K-shape distance")
```

## Cluster analysis
- Applied a clustering algorithm with Dynamic Time Warping (DTW) Distance
- Interpret the spatial distribution of the clusters
- Interpret what case/death growth features/patterns does each cluster represents
- To predict the cluster membership by demographical and political information for each state

```{r US states geographical data scrapping and correlation, echo=FALSE, incude=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(choroplethr)
library(acs)
```

## Clustering with Dynamic Time Warping (DTW) Distance
- We used dtwclust R package
- We first applied clustering validity index to determine the best K size
- The best K = 6
```{r CVI for clustering, echo=FALSE}
print("Partition clustering for cases")
pc_k_case <- tsclust(ts_avg_case_matrix, k = 5L:9L,
                distance = "dtw_basic", centroid = "dba",norm = "L2",
                preproc = zscore,norm = "L2",
                window.size = 20L)

names(pc_k_case) <- paste0("k_", 5L:9L)
sapply(as.list(pc_k_case), cvi, type = "internal")

print("Partition clustering for deaths")
pc_k_death <- tsclust(ts_avg_death_matrix, k = 5L:9L,
                distance = "dtw_basic", centroid = "dba",norm = "L2",
                preproc = zscore,norm = "L2",
                window.size = 20L)

names(pc_k_death) <- paste0("k_", 5L:9L)
sapply(as.list(pc_k_death), cvi, type = "internal")
```
```{r Clustering, include=FALSE}
pc_case <- tsclust(ts_avg_case_matrix, type = "partitional", k = 6L, preproc = zscore,norm = "L2",
              distance = "dtw_basic", centroid = "dba",
              trace = TRUE, seed = 8,
              norm = "L2", window.size = 20L,
              args = tsclust_args(cent = list(trace = TRUE)))

case_cluster <- pc_case@cluster
case_cluster_list <- as.list(setNames(case_cluster, tolower(states_vect)))


pc_death <- tsclust(ts_avg_death_matrix, type = "partitional", k = 6L, preproc = zscore,norm = "L2",
              distance = "dtw_basic", centroid = "dba",
              trace = TRUE, seed = 8,
              norm = "L2", window.size = 20L,
              args = tsclust_args(cent = list(trace = TRUE)))

death_cluster <- pc_death@cluster
death_cluster_list <- as.list(setNames(death_cluster, tolower(states_vect)))

hc_case <- tsclust(ts_avg_case_matrix, type = "hierarchical", k =7, 
              distance = "dtw_basic", trace = TRUE, preproc = zscore,norm = "L2",
              control = hierarchical_control(method = "average"))

hc_death <- tsclust(ts_avg_death_matrix, type = "hierarchical", k = 7, 
              distance = "sbd", trace = TRUE, preproc = zscore,norm = "L2",
              control = hierarchical_control(method = "average"))
pc_case_plot <- plot(pc_case)
pc_death_plot <- plot(pc_death)
hc_case_plot <- plot(hc_case)
hc_death_plot <- plot(hc_death)
```

## Partition Clustering
```{r plot clutering, echo=FALSE}
grid.arrange(pc_case_plot,pc_death_plot,nrow=2)
```

## Hierarchical Clustering
```{r h clustering plot, fig.width=9, fig.height=5, echo=FALSE, message=FALSE, results='hide'}
par(mfrow=c(1,2)) 
plot(hc_case, cex=0.5)
xlab('')
plot(hc_death, cex=0.5)
```

## What does each cluster means?
```{r , include=FALSE}
centroids_plots_case <- vector('list', 6)
for (i in 1:6) {
  temp <- pc_case@centroids[[i]]
  temp_df <- data.frame(t=1:length(temp), cases=temp)
  centroids_plots_case[[i]] <- ggplot(temp_df, aes(x=t, y=cases)) +  
  geom_line(linetype = "solid", color = brewer.pal(n = 7, name = "Set2")[i+1], size=1) +
    ggtitle(paste("Case Cluster", i))
}

centroids_plots_death <- vector('list', 6)
for (i in 1:6) {
  temp <- pc_death@centroids[[i]]
  temp_df <- data.frame(t=1:length(temp), death=temp)
  centroids_plots_death[[i]] <- ggplot(temp_df, aes(x=t, y=death)) +  
  geom_line(linetype = "solid", color = brewer.pal(n = 7, name = "Set2")[i+1], size=1)+
    ggtitle(paste("Death Cluster", i))
}
```
```{r Cluster centroid plotting, echo=FALSE, results='hide', message=FALSE}
grid.arrange(grobs = c(centroids_plots_case, centroids_plots_death), ncol = 3)
```

## Cluster on a US map
```{r State data, include=FALSE}
state <- map_data("state")
setdiff(unique(state$region), names(death_cluster_list))

death_cluster_list['district of columbia'] = 0
case_cluster_list['district of columbia'] = 0

state$case_cluster <- unlist(case_cluster_list[state$region], use.names=FALSE)
state$death_cluster <- unlist(death_cluster_list[state$region], use.names=FALSE)

state$case_cluster <- as.character(state$case_cluster)
state$death_cluster <- as.character(state$death_cluster)
```

```{r plot cluster map,echo=FALSE,fig.width=9, fig.height=4}
us_cluster_case <- ggplot(data=state, aes(x=long, y=lat, fill=case_cluster, group=group)) + 
  geom_polygon(color = "white") + 
  guides(fill=FALSE) + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) +
  scale_fill_manual(values=brewer.pal(n = 7, name = "Set2")) + 
  ggtitle('Clustering of new cases US states') + 
  coord_fixed(1.3)

us_cluster_death <- ggplot(data=state, aes(x=long, y=lat, fill=death_cluster, group=group)) + 
  geom_polygon(color = "white") + 
  guides(fill=FALSE) + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
  scale_fill_manual(values=brewer.pal(n = 7, name = "Set2")) + 
  ggtitle('Clustering of new deaths US states') + 
  coord_fixed(1.3)

grid.arrange(us_cluster_case, us_cluster_death, ncol = 2)
```

## To predict cluster membership
- We fitted a multinomial logistic regression model
  - total population, percent white, median age, per capita income, political party affiliation, and the population density

$$ f(k, i) = \beta_{0, k} + \beta_{1, k} X_{population, i} $$
$$ + \beta_{2, k} X_{white percentage, i} $$
$$ + \beta_{3, k} X_{income, i} + \beta_{4, k} X_{age, i} $$
$$ + \beta_{5, k} X_{party affiliation, i} + \beta_{6, k} X_{population density, i} $$
- We applied a Chi-square test for the performance of the model
- We applied a repeated Likelihood Ratio Test

```{r construct data for multinomial testing, include=FALSE}
us_data_2020 <- read.csv(file = 'us_data_2020.csv')
prediction_clusters_case <- unlist(case_cluster_list[us_data_2020$region], use.names=FALSE)
prediction_clusters_death <- unlist(death_cluster_list[us_data_2020$region], use.names=FALSE)

vote <- c(0, 0, 1, 0, 1, 1, 1, 1, 0, 1, 1, 0, 1, 0,0,0,0,0, 1, 1,1,1,1, 0,0,0,0, 1,1,1,1,1, 0,0,0,0, 1,1,1, 0,0,0,0,0, 1,1,1,0,1,0)

us_data_2020$cluster_case <- prediction_clusters_case
us_data_2020$cluster_death <- prediction_clusters_death
states_vect <- state.name;
us_data_2020_subset <- subset(us_data_2020, region %in% tolower(states_vect))
us_data_2020_subset$vote <- vote

us_data_2020_subset$cluster_case <- as.factor(us_data_2020_subset$cluster_case)
us_data_2020_subset$cluster_death <- as.factor(us_data_2020_subset$cluster_death)
us_data_2020_subset$vote <- as.factor(us_data_2020_subset$vote)

us_data_2020_subset$total_population_scale <- scale(us_data_2020_subset$total_population)
us_data_2020_subset$percent_white_scale <- scale(us_data_2020_subset$percent_white)
us_data_2020_subset$per_capita_income_scale <- scale(us_data_2020_subset$per_capita_income)
us_data_2020_subset$median_rent_scale <- scale(us_data_2020_subset$median_rent)
us_data_2020_subset$median_age_scale <- scale(us_data_2020_subset$median_age)
us_data_2020_subset$density <- scale(state.density)
```

```{r multinomial model fitting, include=FALSE}
multi_mo_case <- multinom(cluster_case ~ total_population_scale + percent_white_scale +
                       per_capita_income_scale + vote + density + 
                       median_age_scale, data = us_data_2020_subset, model=TRUE)
multi_mo_death<- multinom(cluster_death ~ total_population_scale + percent_white_scale +
                       per_capita_income_scale + vote + density + 
                       median_age_scale, data = us_data_2020_subset, model=TRUE)
```

## Model fitting Chi square
```{r chi square testing, echo=FALSE, message=FALSE, warning=FALSE}
print('Chi square testing for the full model, case')
chisq.test(us_data_2020_subset$cluster_case,predict(multi_mo_case))
print('Chi square testing for the full model, death')
chisq.test(us_data_2020_subset$cluster_death,predict(multi_mo_death))
```


## The results for each predictors (Likelihood Ratio Test)
- for cases
  - percentage of white polulation (p < 0.001)
  - the political party affliation (p < 0.05) 
- for deaths
  - the polulation density (p < 0.001)
  - the percentage of white polulation (p = 0.051)


## Conclusion for Spatial Trend Analysis
 - Cases/deaths growth patterns are correlated with their geographical location.
   - Adjacently spreading/difussion
 - 6 clusters for cases/deaths
   - Timing of waves
   - Relative amount of different waves
 - The clusters are also geographically clustered
   - the adjancent spreading of the virus
   - the natural environment of different states (e.g. temperature, weather, etc.) 
   - the way people live/test (e.g., urban vs. rural)

## Conclusion for Spatial Trend Analysis
 - percentage of white polulation and political party affliation predicts cases
   - race inequality in time of pandemic
   - political influence on spread of virus
 - population density is the major predictor for death
   - hospitalization resources are crucial for saving patients' lives
   - cope with COVID-19 by population density

# Thank you